---
title: Global Mortality
author: Phil Roberts
date: '2018-04-16'
slug: global-mortality
categories:
  - Tidy Tuesday
tags:
  - Health
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(tidytuesdayR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
library(tidyr)
library(here)
library(skimr)
library(kableExtra) #https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html
library(stringr)
library(jsonlite) #for continents: https://datahub.io/core/country-codes#r
library(readr)
library(purrr) #to handle multiple input file
```

```{r functions, include=FALSE}
#Functions
save_plot_date <- function(folder = "temp_plots"){
  ggsave(here::here(paste0(folder,"/plot_",stringr::str_replace_all(Sys.time(),":|-| ","_"),".png")))
}
ggplot_print_replace <- function(x, ...) {
  save_plot_date("temp_plots")
  ggplot2:::print.ggplot(x, ...)
}
#Replace ggplot print method
print.ggplot <- ggplot_print_replace
```

# Global Mortality - Causes and Trends

## Our World in Data - rates

#### Background

[Our World in Data](https://ourworldindata.org/what-does-the-world-die-from) asked 'what does the world die from?':

> In 2017, around [56 million](https://ourworldindata.org/grapher/number-of-deaths-by-age-group) people died -- nearly half of these were aged 70 years or older; 27% aged 50-69; 14% aged 15-49; only 1% aged 5-14; and around 10% were children under the age of 5.
>
> The leading global killer in 2017 were cardiovascular diseases (CVD), which refer to a [range of diseases](https://www.who.int/cardiovascular_diseases/about_cvd/en/) that affect the heart and blood vessels...Cardiovascular diseases were responsible for around one-third of all deaths.
>
> [Cancers](https://ourworldindata.org/how-many-people-in-the-world-die-from-cancer)...were the second largest, claiming around 17%, or every sixth death, globally...Approximately 1.8 million newborns still died due to complications at birth. Diarrheal diseases...claimed 1.6 million people in 2017 and...around 620,000...died from malaria in 2017 ... [Road incidents](https://ourworldindata.org/causes-of-death#road-accidents)...claim(ed) 1.2 million in 2017...The number who die from [suicide](https://ourworldindata.org/suicide) is more than double that of [homicide](https://ourworldindata.org/homicides) at a global level.

#### Get data

Load data from github tidy tuesday repo. Use the tidytuesdayR package. I'm not sure of the etiquette here. Seems wrong to download the data every time. However, my goal is to *not* manually download the data. I tried several other approaches, but could not find a better approach. I suppose if I only download a few times it's ok. Also, grab countinent lookup table. Underlying source is <http://ghdx.healthdata.org/> (University of Washington, part of Institute for Health Metrics and Evaluation IHME).

```{r}
# facts
raw <- tt_load("2018-04-16")

pop_file <- 'https://datahub.io/core/population/datapackage.json'
pop_data <- fromJSON(paste(readLines(pop_file, warn=FALSE), collapse=""))
for(i in 1:length(pop_data$resources$datahub$type)){
  if(pop_data$resources$datahub$type[i]=='derived/csv'){
    path_to_file = pop_data$resources$path[i]
    pop <- read.csv(url(path_to_file)) %>%
      dplyr::select(c('country_code'='Country.Code',
                      'year'='Year',
                      'population'='Value')) 
  }
}

# dimensions: comprehensive country codes
ccc_file <- 'https://datahub.io/core/country-codes/datapackage.json'
ccc_data <- fromJSON(paste(readLines(ccc_file, warn=FALSE), collapse=""))
for(i in 1:length(ccc_data$resources$datahub$type)){
  if(ccc_data$resources$datahub$type[i]=='derived/csv'){
    path_to_file = ccc_data$resources$path[i]
    ccc <- read.csv(url(path_to_file)) %>%
      # replace_na(list(Continent_Code='NA')) %>%
      dplyr::select(c('continent'='Continent', 
                      'region'='Region.Name', 
                      'subregion'='Sub.region.Name',
                      # 'country'='official_name_en',
                      'capital'='Capital',
                      'country_code'='ISO3166.1.Alpha.3',
                      'indendent'='is_independent',
                      'development'='Developed...Developing.Countries',
                      'least developed'='Least.Developed.Countries..LDC.')) 
  }
}
```

#### Preview data

Use skimr to skim data. Use kableExtra to make pretty table. I think rowwise() is discouraged, but not sure why. Dataset has a row for each country-year and column for each (n=\~30) disease indicating share of deaths due to the disease.

```{r results='asis'}
df0 <- raw$global_mortality %>%
  left_join(ccc, by=c('country_code'='country_code')) %>%
  left_join(pop, by=c('country_code'='country_code', 'year'='year')) %>%
      replace_na(list(Continent_Code='Smry', Continent_Name='Summary'))

skimmed <- df0 %>% 
  skim() %>%  #histogram workaround from https://github.com/ropensci/skimr/issues/278
  rowwise() %>%
  dplyr::mutate(hist = 
     numeric.hist %>% 
     utf8ToInt() %>% 
     as.character() %>% 
     paste0("&#", . ,";", collapse = "")) %>%
  dplyr::select(-numeric.hist) %>%
  dplyr::rename_with(~ tolower(gsub("skim_|numeric.|character.", "", .x, fixed = FALSE))) %>%
  mutate_if(is.numeric, round, digits=2) %>%
  dplyr::select(c('type','variable','complete_rate','n_unique','hist','mean','sd','min','p25','p50','p75', 'max'))

skimmed %>% 
  kbl(escape = FALSE) %>%
  kable_paper("hover", full_width = F, font_size = 9, fixed_thead = T)
```

\#todo:

-   display row and column count in raw data

-   figure out why n\_unique shows NA

-   shorten 'type'

-   investigate [boxplots built in to kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html#Insert_Images_into_Columns); or sortable tables with [DT](https://rstudio.github.io/DT/)

-   exclude deaths due to natural disaster, conflict, terrorism as these are [volatile](https://ourworldindata.org/is-it-fair-to-compare-terrorism-and-disaster-with-other-causes-of-death)

-   make table shorter

#### Cardiovascular Disease

Dig into CVD. Due to spaces in field names, need back-tic in ggplot function call.

```{r results='asis'}
#todo learn how to pull out 'Cardiovascular diseases (%)' to define once
# perhaps define this whole chunk as a function using tidy evalutation: https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/
dfcvd <- df0 %>%
  dplyr::select(c('continent',
                  'country',
                  'year',
                  'population',
                  'Cardiovascular diseases (%)')) %>%
  mutate('deaths' = round(population * `Cardiovascular diseases (%)` / 100) )

pltcvd <- dfcvd %>%
  filter(!is.na('continent')) %>%
  ggplot(aes(x=year, y =`Cardiovascular diseases (%)`, group=country) ) +
  geom_smooth(aes(y =`Cardiovascular diseases (%)`, color=continent) ) +
  facet_wrap(~ continent, scale = "fixed", shrink=TRUE, ncol = 3, strip.position = 'left') 
#   scale_x_date(date_labels = "%Y")
pltcvd
```

## Numerators and Denominators - GHDx

source: <http://ghdx.healthdata.org/gbd-results-tool>

```{r results='asis'}
IHME0 <-
  list.files(path='../../data/', pattern = 'IHME-GBD_2019_DATA-689c8af3*', full.names = TRUE) %>%
  map_df(~read_csv(.))

IHMEcodes <- read_csv("../../data/IHME_GBD_2019_CODEBOOK_Y2020M11D25.CSV")
IHMEmeasure <- dplyr::select(IHMEcodes,c('measure_id','measure_name')) %>%
  filter(!is.na(as.numeric(measure_id))) %>%
  mutate('measure_id' = as.numeric(measure_id))
IHMElocation <- dplyr::select(IHMEcodes,c('location_id','location_name')) %>%
  filter(!is.na(as.numeric(location_id))) %>%
  mutate('location_id' = as.numeric(location_id))
IHMEsex <- dplyr::select(IHMEcodes,c('sex_id','sex_label')) %>%
  filter(!is.na(as.numeric(sex_id))) %>%
  mutate('sex_id' = as.numeric(sex_id))
IHMEage <- dplyr::select(IHMEcodes,c('age_group_id','age_group_name')) %>%
  filter(!is.na(as.numeric(age_group_id))) %>%
  mutate('age_group_id' = as.numeric(age_group_id))
IHMEcause <- dplyr::select(IHMEcodes,c('cause_id','cause_name')) %>%
  filter(!is.na(as.numeric(cause_id))) %>%
  mutate('cause_id' = as.numeric(cause_id))
# IHMErei <- dplyr::select(IHMEcodes,c('rei_id','rei_name')) %>%
#   filter(!is.na(as.numeric(rei_id))) %>%
#   mutate('rei_id' = as.numeric(rei_id))
IHMEmetric <- dplyr::select(IHMEcodes,c('metric_id','metric_name')) %>%
  filter(!is.na(as.numeric(metric_id))) %>%
  mutate('metric_id' = as.numeric(metric_id))

IHME1 <- IHME0 %>%
  left_join(IHMEmeasure,  by=c('measure' ='measure_id')) %>%
  left_join(IHMElocation, by=c('location'='location_id')) %>%
  left_join(IHMEsex,      by=c('sex'     ='sex_id')) %>%
  left_join(IHMEage,      by=c('age'     ='age_group_id')) %>%
  left_join(IHMEcause,    by=c('cause'   ='cause_id')) %>%
  left_join(IHMEmetric,   by=c('metric'  ='metric_id'))
```

```{r}
IHMEnum <- IHME0 %>%
  filter(metric == 1) %>% #metric_id 1 is number
  dplyr::select(c(
    measure,
    # measure_name,
    location,
    # location_name,
    sex,
    # sex_name,
    age,
    # age_name,
    cause,
    # cause_name,
    year,
    'number'=val
  ))

IHMErate <- IHME0 %>%
  filter(metric == 3) %>% #metric_id 3 is rate
  dplyr::select(c(
    measure,
    location,
    sex,
    age,
    cause,
    year,
    'rate'=val,
    'rate_upper'=upper,
    'rate_lower'=lower
  ))

IHME1 <- IHMEnum %>%
  inner_join(IHMErate, by = c('measure','location','sex','age','cause','year'))

skmIHME <- IHME1 %>% 
  skim() %>%  #histogram workaround from https://github.com/ropensci/skimr/issues/278
  rowwise() %>%
  dplyr::mutate(hist = 
     numeric.hist %>% 
     utf8ToInt() %>% 
     as.character() %>% 
     paste0("&#", . ,";", collapse = "")) %>%
  dplyr::select(-numeric.hist) %>%
  dplyr::rename_with(~ tolower(gsub("skim_|numeric.|character.", "", .x, fixed = FALSE))) %>%
  mutate_if(is.numeric, round, digits=2) #%>%
  # dplyr::select(c('type','variable','complete_rate','n_unique','hist','mean','sd','min','p25','p50','p75', 'max'))

skmIHME %>% 
  kbl(escape = FALSE) %>%
  kable_paper("hover", full_width = F, font_size = 9, fixed_thead = T)
```
